Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-28 submission/manuscript.Rmd
shigeki edits results

As of yet, studies on germline gene silencing mechanisms have focused on a handful of key germline genes. The lack of a germline gene list prohibits a comprehensive, genome-wide analysis of their de-repression and silencing mechanisms. We therefore generated a list of germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013]. We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any non-gonadal tissue of adult wild type mice[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted gonads, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally characterize germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).

@@

We first identified _Kdm5c_-KO EpiLC DEGs in our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and then assessed tissue-enriched gene expression. Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B). Markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_, were unaltered in _Kdm5c_-KO EpiLCs (Figure 3C). Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3D), indicating the misexpression of tissue-enriched genes with KDM5C loss does not impair EpiLC formation.

We next compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development. We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F). EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023]. However, we did not see misexpression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell-like state. Altogether, these results demonstrate that while the _Kdm5c_-KO brain primarily expresses spermiogenesis genes, _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis.

@@

Homozygous and heterozygous _Kdm5c_ knockout females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A). While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs, such as _Tktl2_ and _Esx1_ (Figure 4B). We compared the known functions of germline genes dysregulated only in females (XX only - dysregulated in XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared). Female-specific germline DEGs were enriched for meiotic (GO:0051321 meiotic cell cycle) and flagellar (GO:0003341 cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 protein localization to mitochondrion). Germline transcripts expressed in both sexes were enriched for meiotic (GO:0140013 meiotic nuclear division) and egg-specific functions (GO:0007292 female gamete generation).

The majority of germline genes expressed in both sexes had a greater log2 fold change in females compared to males (Figure 4D). This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated germline gene suppression. Female sensitivity could be due to improper XCI in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, female germline DEGs were not biased towards the X chromosome and had a a similar overall proportion of X chromosome DEGs compared to males (XY _Kdm5c_-KO - 10.29%, XX _Kdm5c_-HET - 7.43%, XX _Kdm5c_-KO - 10.59%) (Figure 4G). For both males and females, the majority of germline DEGs instead lie on autosomes (Figure 4G). Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of potential defects in XCI.

@@

While many germline genes act in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. Therefore, we wondered if _Kdm5c_ mutant males were more likely to express sperm genes and if mutant females would instead express egg genes. To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg and sperm-biased genes. We defined germ cell sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes (Figure 4H). We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-J). Germline genes dysregulated exclusively in _Kdm5c_ mutant males and females were also not biased towards their corresponding germ cell sex (Figure 4I). This indicates sex differences in germline gene dysregulation is not due to sex-specific activation of sperm or egg transcriptional programs. Altogether, these results demonstrate that the sex of _Kdm5c_ mutant cells influences the degree of germline gene, independent of germline gene sex.

- **note: the edited last sentence ended with "independent of germ cell sex", but I think if it's read out of context that sounds like we tested KDM5C in the four core genotypes. But I'm not sure if "germline gene sex" is confusing**

@@

KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but its binding is absent in neurons[@scandagliaLossKdm5cCauses2017]. However, the lack of a comprehensive list of germline-enriched genes prohibited systematic characterization of KDM5C binding at germline gene promoters. Thus, it is unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.

@@

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 5B). Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C). Germline-specific ontologies were only enriched in promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C). There were no ontologies significantly enriched for genes only bound by KDM5C in PNCs. We next evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes. In EpiLCs, we observed modest KDM5C signal at about half of all germline genes (Figure 5D). Based on our ChIP-seq peak cut-off criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 5E). One notable gene that lacked KDM5C binding was _Stra8_, even though its mRNA is expressed in _Kdm5c_-KO EpiLCs (Figure 5F). In EpiLCs, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%), but 3 out of the 4 genes dysregulated in both the brain and EpiLCs (Supplementary figure XXX). In condordance with our gene ontology results, we did not observe KDM5C accumulation at any germline gene promtoers in PNCs (Figure 5D, Supplementary figure XXX). Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including enrichment of meiotic genes, but does not directly regulate germline genes in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their promoters.

>>>>

As of yet, research on germline silencing mechanisms has focused on a handful of key genes rather than assessing germline gene suppression genome-wide due to the lack of a comprehensive mouse germline gene list. We therefore generated a list of germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013]. We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any non-gonadal tissue of adult wild type mice[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted gonads, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally characterize germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).

@@

_Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3A) and EpiLCs properly expressed markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3B). We then identified tissue-enriched DEGs in our previously published RNA-seq dataset of wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1). Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B).

Using the list of mouse germline-enriched genes curated above, we found `r nrow(epigerm)` were misexpressed in male _Kdm5c_-KO EpiLCs. We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development. We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F). EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H). These genes are typically expressed when PGCs commit to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023]. However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell-like state. Altogether, these results demonstrate _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis while the _Kdm5c_-KO brain primarily expresses spermiogenesis genes.

@@

Homozygous and heterozygous _Kdm5c_ knockout females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A). While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs, such as _Tktl2_ and _Esx1_ (Figure 4B). We compared the known functions of germline genes dysregulated only in females (XX only - dysregulated in XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared) (Figure 4C). Female-specific germline DEGs were enriched for meiotic (GO:0051321 meiotic cell cycle) and flagellar (GO:0003341 cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 protein localization to mitochondrion). Germline transcripts expressed in both sexes were enriched for meiotic (GO:0140013 meiotic nuclear division) and egg-specific functions (GO:0007292 female gamete generation).

The majority of germline genes expressed in both sexes had a greater log2 fold change in females compared to males (Figure 4D). This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated germline gene suppression. Female sensitivity could be due to impaired XCI in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, female germline DEGs were not biased towards the X chromosome and had a a similar overall proportion of X chromosome DEGs compared to males (XY _Kdm5c_-KO - 10.29%, XX _Kdm5c_-HET - 7.43%, XX _Kdm5c_-KO - 10.59%) (Figure 4G). The majority of germline DEGs instead lie on autosomes for both male and female _Kdm5c_ mutants (Figure 4G). Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of XCI defects.

@@

While many germline genes act in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. Therefore, we wondered if _Kdm5c_ mutant males were more likely to express sperm genes and if mutant females would instead express egg genes. To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg and sperm-biased genes. We defined germ cell sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes (Figure 4H). We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-J). Germline genes dysregulated exclusively in _Kdm5c_ mutant males and females were also not biased towards their corresponding germ cell sex (Figure 4I). This indicates sex differences in germline gene dysregulation is not due to sex-specific activation of sperm or egg transcriptional programs. Altogether, these results demonstrate that both male and female _Kdm5c_ mutant cells express both sperm and egg-biased genes.

@@

KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but its binding is absent in neurons[@scandagliaLossKdm5cCauses2017]. However, the lack of a comprehensive list of germline-enriched genes prohibited comprehensive characterization of KDM5C binding at germline gene promoters. Thus, it is unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.

@@

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 5B). Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C). Germline-specific ontologies were only enriched in KDM5C-bound promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C). There were no ontologies significantly enriched for genes bound by KDM5C only in PNCs. We next evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes. KDM5C binding around the TSS of many germline genes is evident in EpiLCs, but absent in PNCs (Figure 5D). Based on our ChIP-seq peak cut-off criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 5E).

 In EpiLCs, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%). However, KDM5C was bound to the promoter at 3 out of the 4 genes dysregulated in both the brain and EpiLCs (Supplementary figure XXX). Representative examples of KDM5C-bound and unbound germline DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F). Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including enrichment of meiotic genes, but does not directly regulate germline genes in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their promoters.

 One notable gene that lacked KDM5C binding was _Stra8_, even though its mRNA is expressed in _Kdm5c_-KO EpiLCs ().
